Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2012-02-16 boettiger.tex
minor edits and touchups
spellcheck and ready to go

%\documentclass[authoryear,5p]{elsarticle}
%\documentclass[authoryear,preprint,11pt]{elsarticle}
\documentclass[authoryear,review,11pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
\linenumbers
\usepackage[pdftex, colorlinks]{hyperref}
\usepackage{float}

@@

\journal{\tiny Journal of the Royal Society Interface}
\begin{document}
\begin{frontmatter}
  \title{Quantifying Limits to Detection of Early Warning for Critical Transitions}
  \author[cpb]{Carl Boettiger\corref{cor1}}
  \ead{cboettig@ucdavis.edu}
  \author[esp]{Alan Hastings}
  %\cortext[cor1]{Corresponding author.}
  \address[cpb]{Center for Population Biology, 1 Shields Avenue, University of California, Davis, CA, 95616 United States.}
  \address[esp]{Department of Environmental Science and Policy, University of California, Davis}

@@

%%% hmm, what about all the red-herring topics, like change point analysis?  Add a discussion such as Contamin & Ellison ~\citep{Contamin2009} have about forecasting shifts, not post-hoc identification of shifts.

@@

\subsubsection*{Problematic null models}
Specifying an appropriate null model is also difficult.
Non-parametric null hypotheses seem to require the fewest assumptions but in fact can be the most problematic.
For instance, the standard non-parametric hypothesis test with Kendall's tau rank correlation coefficient
assumes only that the two variables are independent,
but this is an assumption that is violated by the very experimental design:
temporal correlations will exist in any finely-enough sampled time series,
and moving windows introduce temporal correlations in the statistics.
Under such a test any adequately large data set will find a significant result,
regardless of whether a warning signal exists.
A similar problem arises when the points in the time series are reordered to create a null hypothesis --
this destroys the natural autocorrelation in the time series.
More promising parametric null models have been proposed,
such as autoregressive models in~\citet{Dakos2008}, bringing us
closer to a model-based approach with explicit assumptions.
Others have looked for alternative summary statistics where
reasonable null models are more readily available,
such as~\citet{Seekell2011}'s proposal to test for conditional heteroscedaticity.

@@

where $P(n,t)$ is the probability of having $n$ individuals at time $t$, $b_n$ is the probability of a birth event occuring in a population of $n$ individuals, $d_n$ the probability of a death.  $e, K, h$ and $a_t$ are parameters.  
This corresponds to the well-studied ecosystem model of over-exploitation~\citep{Noy-Meir1975, May1977},
with stochasticity introduced directly through the demographic process.
We select this model since it is has discrete numbers of individuals, 
nonlinear processes, and the noise is driven by Poisson process of births and deaths instead of a Gaussian, 
and thus provides an illustration that our approach is robust to the violations of those
assumptions in model~\eqref{LSN}.

@@

The use of box plots in Figure~\ref{fig:summary_box} provide a convient and familiar way 
to visualize the overlap between more than two distributions,
though they lack the resolution of the overlapping density distributions in 
Figures~\ref{fig:simulation},~\ref{fig:chemostat}, \ref{fig:glaciation}.  
The overlapping distributions are the natural representation from which to introduce the ROC curve, as in Figure~\ref{fig:roc_intro}.

@@

 The ROC curves for these data (Fig.~\ref{fig:summary_roc}) show that the summary-statistic based indicators
frequently lack the sensitivity to distinguish reliably between observed patterns from a stable or unstable system.
The large correlations observed in the empirical examples (Fig.~\ref{fig:summary_stats}) are not uncommon in stable systems.
It is notable that in both empirical examples the summary statistics approach does little better than chance in distinguishing 
replicates that have been simulated from models~\eqref{LSN} and ~\eqref{OU}, 
despite the fact that these models correspond 
to the assumptions of the summary statistics approaches.  
On the simulated data, the varianced based method approaches the 
true-positive rate of our likelihood method at higher levels of false positives, 
but performs worse when the desired level of false postives is low.  
The ROC curve helps us compare the performance of the different approaches at different tolerances.  
For instance, Table~\ref{tab:falsepos} shows the fraction of true crashes caught at a 5\% false positive rate.
We can instead set a desired True positive rate and read off the resulting number of false alarms, Table~\ref{tab:truepos}.

@@

 % past tense or not?
We have used three examples to illustrate the performance of this approach in data from 
simulation, a chemostat experiment, and paleo-atmospheric record; examples differing in 
sampling intensity and strength of signal of an approaching collapse.  
While the well-sampled geological data shows an unmistakable signal in this model-based approach,
the uncertainty in the smaller simulated and experimental data forces a trade-off between errors.  
%Because such uncertainty may be more typical of early warning signal detection 
%than the monotonically increasing patterns often used to illustrate the approach,

As a way to clearly illustrate the choices involved in looking for warning signals while avoiding false alarms,
we introduce an approach based on receiver operator curves. 
These curves illustrate the extent to which an potential warning signal mitigates the trade-off 
between missed events and false alarms.  
The extent of the difficulty in finding reliable indicators of impending regime shifts
based on summary statistics becomes clear from the ROC curves of these statistics,
where a 5\% false positive rate often corresponds to only a 5\% true positive rate,
performing no better than the flip of a coin. %%%% EEEK can we really say it that strongly?
By estimating the ROC curve for a given set of data,
we can better avoid applying warning signals in cases of inadequate power.
By taking advantage of the assumptions being made to write down a specific likelihood function,
we can develop approaches that get the most information from the data available.

%%% Model adequacy
In any application of early warning signals, it is essential to address the question of model adequacy.
Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.
As the bifurcation results from the principle eigenvalue passing through zero,
the warning signal is expected in linear-order dynamics;
estimation of the nonlinear model is less powerful and less accurate.
The performance of this approach in the simulated data -- which is nonlinear in its dynamics
and driven with non-Gaussian noise introduced by the Poisson demographic events --
demonstrates the accuracy under violation of these assumptions.
% this robustness to assumptions can also be seen in the performance of the empirical systems

>>>>

%\documentclass[authoryear,5p]{elsarticle}
%\documentclass[authoryear,preprint,11pt]{elsarticle}
\documentclass[authoryear,review,11pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
\linenumbers
\usepackage[pdftex, colorlinks]{hyperref}
%\usepackage{float}
%\usepackage{endfloat}

@@

\journal{\tiny Journal of the Royal Society Interface}
\begin{document}
\begin{frontmatter}
  \title{Quantifying Limits to Detection of Early Warning for Critical Transitions}
  \author[cpb]{Carl Boettiger\corref{cor1}}
  \ead{cboettig@ucdavis.edu}
  \author[esp]{Alan Hastings}
  %\cortext[cor1]{Corresponding author.}
  \address[cpb]{Center for Population Biology, 1 Shields Avenue, University of California, Davis, CA, 95616 United States.}
  \address[esp]{Department of Environmental Science and Policy, University of California, Davis, CA, 95616 United States}

@@

@@

\subsubsection*{Problematic null models}
Specifying an appropriate null model is also difficult.
Non-parametric null hypotheses seem to require the fewest assumptions but in fact can be the most problematic.
For instance, the standard non-parametric hypothesis test with Kendall's tau rank correlation coefficient
assumes only that the two variables are independent,
but this is an assumption that is violated by the very experimental design:
temporal correlations will exist in any finely-enough sampled time series,
and moving windows introduce temporal correlations in the statistics.
Under such a test any adequately large data set will find a significant result,
regardless of whether a warning signal exists.
A similar problem arises when the points in the time series are reordered to create a null hypothesis --
this destroys the natural autocorrelation in the time series.
More promising parametric null models have been proposed,
such as autoregressive models in~\citet{Dakos2008}, bringing us
closer to a model-based approach with explicit assumptions.
Others have looked for alternative summary statistics where
reasonable null models are more readily available,
such as~\citet{Seekell2011}'s proposal to test for conditional heteroscedasticity.

@@

where $P(n,t)$ is the probability of having $n$ individuals at time $t$, $b_n$ is the probability of a birth event occurring in a population of $n$ individuals, $d_n$ the probability of a death.  $e, K, h$ and $a_t$ are parameters.  
This corresponds to the well-studied ecosystem model of over-exploitation~\citep{Noy-Meir1975, May1977},
with stochasticity introduced directly through the demographic process.
We select this model since it is has discrete numbers of individuals, 
nonlinear processes, and the noise is driven by Poisson process of births and deaths instead of a Gaussian, 
and thus provides an illustration that our approach is robust to the violations of those
assumptions in model~\eqref{LSN}.

@@

The use of box plots in Figure~\ref{fig:summary_box} provide a convenient and familiar way 
to visualize the overlap between more than two distributions,
though they lack the resolution of the overlapping density distributions in 
Figures~\ref{fig:simulation},~\ref{fig:chemostat}, \ref{fig:glaciation}.  
The overlapping distributions are the natural representation from which to introduce the ROC curve, as in Figure~\ref{fig:roc_intro}.

@@

 The ROC curves for these data (Fig.~\ref{fig:summary_roc}) show that the summary-statistic based indicators
frequently lack the sensitivity to distinguish reliably between observed patterns from a stable or unstable system.
The large correlations observed in the empirical examples (Fig.~\ref{fig:summary_stats}) are not uncommon in stable systems.
It is notable that in both empirical examples the summary statistics approach does little better than chance in distinguishing 
replicates that have been simulated from models~\eqref{LSN} and ~\eqref{OU}, 
despite the fact that these models correspond 
to the assumptions of the summary statistics approaches.  
On the simulated data, the variance based method approaches the 
true-positive rate of our likelihood method at higher levels of false positives, 
but performs worse when the desired level of false positives is low.  
The ROC curve helps us compare the performance of the different approaches at different tolerances.  
For instance, Table~\ref{tab:falsepos} shows the fraction of true crashes caught at a 5\% false positive rate.
We can instead set a desired True positive rate and read off the resulting number of false alarms, Table~\ref{tab:truepos}.

@@

We have used three examples to illustrate the performance of this approach in data from 
simulation, a chemostat experiment, and paleo-atmospheric record; examples differing in 
sampling intensity and strength of signal of an approaching collapse.  
While the well-sampled geological data shows an unmistakable signal in this model-based approach,
the uncertainty in the smaller simulated and experimental data forces a trade-off between errors.

As a way to clearly illustrate the choices involved in looking for warning signals while avoiding false alarms,
we introduce an approach based on receiver operator curves. 
These curves illustrate the extent to which an potential warning signal mitigates the trade-off 
between missed events and false alarms.  
The extent of the difficulty in finding reliable indicators of impending regime shifts
based on summary statistics becomes clear from the ROC curves of these statistics,
where a 5\% false positive rate often corresponds to only a 5\% true positive rate,
performing no better than the flip of a coin. 
By estimating the ROC curve for a given set of data,
we can better avoid applying warning signals in cases of inadequate power.
By taking advantage of the assumptions being made to write down a specific likelihood function,
we can develop approaches that get the most information from the data available.

In any application of early warning signals, it is essential to address the question of model adequacy.
Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.
As the bifurcation results from the principle eigenvalue passing through zero,
the warning signal is expected in linear-order dynamics;
estimation of the nonlinear model is less powerful and less accurate.
The performance of this approach in the simulated data -- which is nonlinear in its dynamics
and driven with non-Gaussian noise introduced by the Poisson demographic events --
demonstrates the accuracy under violation of these assumptions.
% this robustness to assumptions can also be seen in the performance of the empirical systems
